Analyzing Strawberry Pricing Strategies Across Canadian Vendors

### 2024-12-03 paper/paper.qmd
finish the model part

---
title: "Analyzing Strawberry Pricing Strategies Across Canadian Vendors"
subtitle: "A Data-Driven Investigation of Seasonal Trends, Vendor Variability, and Promotional Impacts on Grocery Pricing"
author: 
  - Tianning He
thanks: "Code and data are available at: https://github.com/HeTianning/The-price-of-strawberries"
date: 25 November 2024
date-format: long
abstract: "This study investigates the pricing of strawberries across eight major suppliers in Canada: Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods. Using a dataset containing variables such as current price, old price, and month, a Bayesian regression model was applied to analyze the relationships between pricing strategies, seasonal trends, and vendor-specific practices. Results reveal significant price variation among suppliers, with old price showing a strong positive correlation with current price, reflecting the impact of prior pricing strategies on current prices. Seasonal trends captured by the month variable indicate subtle but consistent fluctuations in pricing, highlighting the influence of supply and demand dynamics. These findings provide actionable insights into supplier-specific pricing strategies and market behavior, offering valuable guidance to consumers and stakeholders in the Canadian strawberry market."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(tinytex)
library(knitr)
library(ggplot2)
library(dplyr)
library(arrow)
library(here)
library(modelsummary)
library(bayesplot)
library(rstanarm)
```

@@

# Data {#sec-data}
To simulate, test, download, clean, and model data throughout this analysis, the statistical programming language R was used [@citeR]. Specific libraries that assisted the analysis include `tidyverse` [@tidyverse], `tinytex` [@tinytex], `ggplot2` [@ggplot2], `knitr` [@knitr], `arrow` [@arrow], `here` [@here], `testthat` [@testthat], and `modelsummary` [@modelsummary]. 
## Overview
The data for this analysis was provided by Project Hammer [@filipp2024hammer], which was collected through a process of scraping website UI screens from vendors within the Toronto community. It contains data on product details (e.g., product name, brand, current and old price, and unit of product) for 8 different grocery stores (Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart Canada, and Save-On-Foods). Project Hammer's data is available from February 28, 2024, and since the data is collected by extracting information from grocery store websites, there may be missing data, as discussed in appendix. Through the method of screen scraping, the raw data is intended to measure and reflect grocery store vendor decisions and product pricing trends.
For this analysis, the eight vendors - Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart Canada, and Save-On-Foods - and their pricing for strawberries were considered. The variable “price” for each supplier in this analysis reflects the current price of their product at the time the data was downloaded.

@@

The dataset captures strawberry pricing using structured variables that provide a comprehensive basis for analyzing supplier-specific pricing strategies and seasonal trends. The primary variable of interest is current price, a continuous measure representing the current price of strawberries in Canadian dollars. This variable serves as the dependent variable in the analysis and reflects the vendors’ updated or promotional pricing.

An additional variable, old price, is another continuous measure representing the original price of strawberries before discounts or promotional adjustments. The difference between old price and current price can be calculated to assess the magnitude of price reductions, providing insights into sales promotions, markdown strategies, or vendor-specific pricing policies. Both current price and old price are essential for understanding vendor behavior and the dynamics of pricing over time.

The dataset also includes vendor, a categorical variable identifying one of eight suppliers (Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods). This variable is crucial for comparing pricing strategies across vendors and identifying patterns or outliers within the market.

The month variable is a numeric measure that records the month of observation, enabling the analysis of seasonal trends and price fluctuations over time. By combining month with pricing data, the dataset facilitates the exploration of how factors like demand, supply, and promotions vary throughout the year.

Together, these variables enable a granular and aggregate analysis of strawberry pricing in the Canadian market. The dataset allows us to examine the relationship between current and old prices, evaluate vendor-specific strategies, and uncover seasonal patterns, providing actionable insights for both consumers and suppliers.

```{r}
#| echo: false
#| message: false
#| label: tbl-data
#| tbl-cap: "Sample of the Dataset" 
# Display the first 10 rows of the dataset in a formatted table
col_names <- colnames(analysis_data)

# Display the first 10 rows of the dataset in a formatted table
analysis_data %>%
  head(10) %>%  # Select the first 10 rows
  kable(
    col.names = col_names, # Dynamically use the dataset's column names
    caption = "Table 1: Sample of the Dataset",
    digits = 2, # Format numeric columns to 2 decimal places
    format = "html", # Output format
    align = c("l", "r", "r", "c") # Align columns
  )

## Outcome variables

@@

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-currentpricebyvendor
#| fig-cap: Average Current Price by Vendor
#| echo: false
ggplot(analysis_data, aes(x = vendor, y = current_price, fill = vendor)) +
  stat_summary(fun = "mean", geom = "bar", color = "black") +
  labs(
    x = "Vendor",
    y = "Average Current Price (CAD)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

```{r}
#| label: fig-comparisonofprice
#| fig-cap: Comparison of Current and Old Prices by Vendor
#| echo: false
#| warning: false
#| message: false
ggplot(analysis_data_long, aes(x = price_type, y = price, fill = price_type)) +
  geom_boxplot() +
  facet_wrap(~ vendor) +
  labs(
    x = "Price Type",
    y = "Price (CAD)"
  ) +
  theme_minimal()

The graph(@fig-comparisonofprice) compares current and old strawberry prices across vendors, revealing significant price variability for `Metro` and `Walmart`, likely due to a diverse product range including premium options, while vendors like `No Frills` and `Save-On-Foods` maintain tight and predictable pricing, reflecting budget-friendly strategies. Outliers, particularly for `Metro` and `Walmart`, suggest the presence of specialty or organic products. The small differences between current and old prices across most vendors indicate either stable pricing strategies or limited promotional activity. These trends highlight distinct market positioning, with budget vendors appealing to cost-conscious shoppers and premium vendors catering to a broader or quality-focused market.

```{r}
#| label: fig-monthlytrend
#| fig-cap: Monthly Trend of Current Prices by Vendor
#| echo: false
#| warning: false
#| message: false
ggplot(analysis_data, aes(x = month, y = current_price, group = vendor, color = vendor)) +
  stat_summary(fun = "mean", geom = "line", size = 1) +
  geom_point(stat = "summary", fun = "mean", size = 2) +
  labs(
    x = "Month",
    y = "Average Current Price (CAD)"
  ) +
  theme_minimal()

```
This graph(@fig-monthlytrend) shows the monthly trend of average current strawberry prices across different vendors over several months. `Metro` consistently exhibits the highest prices, with noticeable fluctuations, peaking in mid-year, indicating potential seasonal effects or changes in supply and demand. In contrast, budget-friendly vendors like `No Frills` and `Save-On-Foods` maintain the lowest prices, demonstrating pricing stability aimed at cost-conscious consumers. Mid-range vendors such as `Loblaws`, `T&T`, and `Voila` show relatively stable but slightly variable trends, with occasional peaks suggesting targeted pricing adjustments. Overall, the graph highlights clear segmentation among vendors, with premium vendors like `Metro` adapting dynamically to market conditions, while budget vendors maintain consistent affordability. This trend could reflect differing business strategies to cater to diverse consumer preferences.

@@

Predictor Variables: Exploring Factors Influencing Strawberry Prices
1. Vendor (vendor)
The vendor variable is categorical, identifying the supplier of strawberries. Each vendor employs unique pricing strategies, leading to significant variability in pricing across the eight vendors (Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods). Premium vendors like Metro and Galleria tend to target quality-conscious consumers, reflected in their higher average prices. In contrast, budget-oriented vendors like No Frills and Save-On-Foods maintain lower prices to attract cost-sensitive shoppers. The variability in pricing, shown through boxplots, highlights how vendor identity influences price distributions and captures the strategic segmentation of the Canadian strawberry market.

2. Old and Current Prices
The dataset includes two continuous price variables: old price (original price) and current price (price after any promotions or discounts). The difference between these two variables provides insights into promotional behavior and markdown strategies. Vendors like No Frills and Save-On-Foods typically show small gaps between old price and current price, suggesting limited use of deep discounts, whereas Metro and Walmart exhibit larger gaps, reflecting more aggressive promotional activity. Boxplots comparing old price and current price by vendor demonstrate how discounts influence consumer pricing perceptions and vendor competition.

3. Month (month)
The month variable is numeric, capturing the month of observation. Seasonal patterns play a crucial role in strawberry pricing, with supply and demand fluctuations driving price variability. For instance, prices may drop during peak harvest months and increase during off-season periods when strawberries are scarce. Line charts tracking average monthly prices show clear temporal trends, with peaks and dips aligning with these seasonal dynamics. By analyzing month in combination with other predictors, this variable helps uncover how vendors adjust pricing strategies to reflect seasonal demand and supply conditions.

4. Combined Factors: Vendor, Price, and Seasonality
The interplay of vendor, old price, current price, and month allows for a more comprehensive analysis of pricing strategies. For example, vendors like Metro may use larger discounts (old price vs. current price) during peak seasons to maintain competitiveness, while budget vendors like No Frills may maintain consistent pricing throughout the year. A combined analysis using bar charts and line plots highlights how vendor strategies vary with time, providing insights into the dynamic nature of strawberry pricing.

@@

- Quantify the impact of each predictor (e.g., vendor, original price, and month) on strawberry prices.
- Analyze how the gap between original (`old_price`) and discounted (`current_price`) prices varies across vendors.
- Examine seasonal trends in strawberry prices to identify periods of high or low prices.
- Provide actionable insights into vendor pricing strategies and market dynamics.
## Model Description

@@

### Predictors in the Model

- **Vendor (`vendor`)**: A categorical variable identifying the supplier of strawberries. Different vendors adopt distinct pricing strategies that influence strawberry prices.
- **Original Price (`old_price`)**: A continuous variable representing the original price of strawberries before discounts or promotions. This variable reflects vendor-specific baseline pricing strategies.
- **Month (`month`)**: A numeric variable capturing the observation month, enabling the analysis of seasonal trends and time-based fluctuations.
- **Price Gap (`old_price - current_price`)**: A derived variable quantifying the magnitude of discounts or markdowns applied by vendors.

## Model set-up

@@

**Where**:
- \( y_i \): Current price (`current_price`) of strawberries for observation \( i \) (dependent variable).
- \( \beta_0 \): Intercept, representing the baseline price (reference vendor, average original price, and earliest month).
- \( \beta_{\text{vendor}[i]} \): Vendor-specific effect, capturing how each supplier’s prices deviate from the baseline vendor.
- \( \beta_{\text{old\_price}} \): Effect of the original price, representing how the initial pricing influences the current price.
- \( \beta_{\text{month}} \): Effect of the observation month, capturing seasonal or temporal trends in strawberry pricing.
- \( \epsilon_i \): Residual error, accounting for variability not explained by the predictors.

## Priors in the Bayesian Framework

### Coefficients (\( \beta \)):

@@

This reflects the belief that most predictor effects are small and centered around zero, with moderate variability. For example:
- \(\beta_{\text{old\_price}}\) is expected to have a positive value since higher original prices generally correlate with higher current prices.
- \(\beta_{\text{month}}\) is expected to capture small but meaningful fluctuations due to seasonal trends.

### Residual Standard Deviation (\( \sigma \)):

@@

This prior ensures positivity and reflects typical variance patterns in continuous outcome data.

@@

### Model justification

#### Why Use a Bayesian Linear Regression Model?
A Bayesian linear regression model is a suitable approach for this analysis because it balances simplicity, interpretability, and flexibility while quantifying uncertainty in parameter estimates. This model is particularly effective for exploring the continuous outcome variable, current_price, and its relationship with predictors such as vendor, month, and original price (old price). Unlike traditional linear regression, which assumes a deterministic relationship between predictors and the outcome, the Bayesian framework provides posterior distributions for all parameters, enabling richer and more nuanced inferences. This approach allows us to incorporate prior knowledge, account for uncertainty, and provide probabilistic insights into strawberry pricing.

The Bayesian framework also handles sparse or imbalanced data effectively by leveraging prior distributions to stabilize parameter estimates. For example, when certain vendors have fewer observations, the priors prevent overfitting and provide regularized estimates, making the model robust to variability in data distribution. This ensures reliable insights into vendor-specific pricing strategies, seasonal trends, and the influence of baseline (old price) on observed (current price) prices.
#### Benefits of Using Bayesian Regression

1. **Uncertainty Quantification**:
   - The Bayesian approach provides posterior distributions for each parameter, allowing for uncertainty quantification. For example, the posterior distribution of \( \beta_{\text{vendor}} \) helps quantify the uncertainty around vendor-specific pricing effects.

2. **Incorporating Prior Knowledge**:
   - Priors enable the incorporation of domain knowledge into the analysis. For instance:
     - \( \beta_{\text{vendor}} \): Reflects the expectation that most vendor effects are close to zero, with some deviations.
     - \( \beta_{\text{month}} \): Captures small but meaningful seasonal effects.
     - \( \beta_{\text{old\_price}} \): Represents the assumption that higher original prices lead to higher observed prices.

3. **Handling Sparse Data**:
   - Bayesian models stabilize estimates when some categories (e.g., certain vendors) have fewer observations. Priors regularize the estimates, reducing the risk of overfitting.

4. **Probabilistic Insights**:
   - Posterior distributions provide credible intervals, making it easier to interpret parameter estimates. For instance, a 95% credible interval for \( \beta_{\text{old\_price}} \) of \([0.8, 1.2]\) means there is a 95% probability that the original price effect lies within this range.

#### Explanation of Priors

##### Priors for Coefficients (\( \beta \)):

- **Reasoning**: The normal prior reflects the belief that most predictor effects are small and centered around zero, with moderate variability. For example:
  - \( \beta_{\text{vendor}} \): Reflects the expected variability in pricing strategies across vendors, assuming deviations from the baseline vendor are likely small but not negligible.
  - \( \beta_{\text{old\_price}} \): Expected to be positive, as higher original prices generally result in higher current prices, with moderate variability in the relationship.
  - \( \beta_{\text{month}} \): Captures seasonal effects on pricing, which are typically small but non-zero.

##### Prior for the Intercept (\( \beta_0 \)):

\[
\beta_0 \sim \text{Normal}(0, 2.5)
\]

- **Reasoning**: The intercept represents the baseline price when all predictors are at their reference levels. This prior assumes the baseline price is near zero but allows for moderate deviations.

##### Prior for Residual Standard Deviation (\( \sigma \)):

- **Reasoning**: The exponential prior ensures positivity and reflects typical patterns of variability in continuous outcomes, such as prices. This helps the model account for unobserved variability not explained by the predictors.

#### Choice of Predictors

##### Vendor (`vendor`)

- **Role**: The vendor variable is categorical, capturing supplier-specific pricing strategies. Each vendor adopts distinct pricing approaches based on branding, market positioning, and target demographics.
- **Importance**: By including this variable, the model quantifies how much each vendor's prices deviate from the baseline vendor, providing insights into vendor-specific strategies and market segmentation.

##### Month (`month`)

- **Role**: The numeric `month` variable captures seasonal trends in strawberry pricing. This allows the model to account for temporal effects, such as higher prices during off-seasons or lower prices during harvest periods.
- **Importance**: Including `month` helps the model identify recurring seasonal patterns, enabling a better understanding of time-based price fluctuations.

##### Original Price (`old_price`)

- **Role**: The continuous variable `old_price` represents the baseline price before any discounts or promotions. This predictor captures the relationship between promotional activity and observed prices.
- **Importance**: Including this variable enables the model to assess how the gap between `old_price` and `current_price` varies across vendors and time, providing insights into promotional strategies and pricing adjustments.

@@

```{r}
#| label: fig-CI
#| fig-cap: Posterior Credible Intervals for Model Parameters
#| echo: false
#| warning: false
#| message: false
# Extract posterior draws
posterior_draws <- as.matrix(vendor_price_model)

# Generate credible interval plot for all parameters
mcmc_intervals(
  posterior_draws,
  pars = c("(Intercept)", "vendorLoblaws", "vendorMetro", "vendorNoFrills", 
           "vendorSaveOnFoods", "vendorTandT", "vendorVoila", "vendorWalmart",
           "month", "old_price"),
  prob = 0.95,  # 95% credible interval
  prob_outer = 0.95  # 95% outer credible interval
) +
  xlab("Parameter Estimate")
```
In this graph(@fig-CI), the posterior credible intervals for the Bayesian linear regression model highlight significant differences in strawberry pricing across vendors, seasonal trends, and the influence of original prices. Metro stands out with a clear positive effect, indicating its significantly higher prices compared to the reference vendor, consistent with its premium market positioning. In contrast, budget vendors like Save-On-Foods and No Frills exhibit negative effects, suggesting lower pricing strategies, while other vendors show smaller and more uncertain deviations. The positive, albeit modest, effect of `month` suggests a slight upward trend in prices over time, likely due to seasonal demand or inflation. The strong and precise relationship between `old_price` and `current_price` underscores the proportionality of promotional pricing strategies, where discounts often maintain the hierarchy of original prices. Together, these findings reveal actionable insights into vendor-specific pricing strategies, temporal trends, and the importance of original prices in shaping strawberry pricing dynamics.

```{r}
#| label: fig-posterior
#| fig-cap: Posterior Distribution Histograms for Vendors and Other Parameters
#| echo: false
#| warning: false
#| message: false
# Extract posterior samples
posterior_draws <- as.matrix(vendor_price_model)

>>>>

---
title: "Analyzing Strawberry Pricing Strategies Across Canadian Vendors"
subtitle: "Price Variations Driven by Supplier Strategies and Seasonal Demand in Canadian Grocery Markets"
author: 
  - Tianning He
thanks: "Code and data are available at: https://github.com/HeTianning/The-price-of-strawberries"
date: 25 November 2024
date-format: long
abstract: "This study investigates the pricing of strawberries across eight major suppliers in Canada: Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods. Using a dataset containing variables such as current price, old price, and month, a Bayesian regression model was applied to analyze the relationships between pricing strategies, seasonal trends, and vendor-specific practices. Results reveal significant price variation among suppliers, with old price showing a strong positive correlation with current price, reflecting the impact of prior pricing strategies on current prices. Seasonal trends captured by the month variable indicate subtle but consistent fluctuations in pricing, highlighting the influence of supply and demand dynamics. These findings provide actionable insights into supplier-specific pricing strategies and market behavior, offering valuable guidance to consumers and stakeholders in the Canadian strawberry market."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(knitr)
library(ggplot2)
library(dplyr)
library(arrow)
library(here)
library(modelsummary)
library(bayesplot)
library(rstanarm)
```

@@

# Data {#sec-data}

To conduct the analysis, `R`, a versatile statistical programming language, was utilized for simulating, testing, downloading, cleaning, and modeling data [@citeR]. Several powerful libraries supported various stages of the analysis. The `tidyverse` package suite [@tidyverse], which includes `dplyr` for data manipulation and `ggplot2` for visualization, was central to handling and exploring the dataset. Data cleaning and structuring were enhanced using `janitor`, while `arrow` facilitated efficient handling of large datasets by enabling seamless reading and writing of data [@arrow]. The `here` package streamlined file management, ensuring robust and reproducible workflows [@here].

Visualization and reporting were further supported by `knitr` [@knitr], which enabled dynamic report generation, and `modelsummary` [@modelsummary], which provided professional-quality summaries for statistical models. Bayesian modeling and advanced statistical analysis were conducted using `rstanarm` [@rstanarm], while posterior diagnostics and visualizations were enhanced with `bayesplot` [@bayesplot]. Testing and quality assurance of code were carried out with `testthat` [@testthat], ensuring the reliability and reproducibility of results.

By integrating these libraries the analysis leveraged a comprehensive ecosystem of tools, demonstrating the strength and flexibility of R in handling complex data-driven research workflows.

## Overview

The data for this analysis was provided by Project Hammer [@filipp2024hammer], which was collected through a process of scraping website UI screens from vendors within the Toronto community. It contains data on product details (e.g., product name, brand, current and old price, and unit of product) for 8 different grocery stores (Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart Canada, and Save-On-Foods). Project Hammer's data is available from February 28, 2024, and since the data is collected by extracting information from grocery store websites, there may be missing data, as discussed in appendix. Through the method of screen scraping, the raw data is intended to measure and reflect grocery store vendor decisions and product pricing trends.

For this analysis, the eight vendors - Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart Canada, and Save-On-Foods - and their pricing for strawberries were considered. The variable “price” for each supplier in this analysis reflects the current price of their product at the time the data was downloaded.

@@

The dataset provides a detailed and structured framework for analyzing strawberry pricing dynamics in the Canadian market, offering critical insights into supplier-specific pricing strategies and seasonal trends. The central variable of interest is the current price, a continuous measure representing the price of strawberries in Canadian dollars at the time of observation. This variable acts as the dependent variable in the analysis and reflects real-time vendor pricing decisions, including adjustments for promotions, sales, or other market-driven factors. The current price is essential for understanding how vendors position their products in response to competition and consumer demand.

Complementing this, the dataset includes the old price, another continuous variable that captures the price of strawberries before any discounts or promotional offers. The old price serves as a benchmark for evaluating vendor markdowns, sales strategies, and pricing flexibility. By calculating the difference between the old price and the current price, the analysis can quantify the extent and frequency of price reductions, revealing patterns in promotional behavior. For instance, larger discounts may indicate aggressive marketing tactics during periods of high supply or low demand. Together, current and old prices allow for a nuanced exploration of how pricing evolves over time and in response to market conditions.

The dataset also incorporates vendor as a categorical variable, identifying one of eight major suppliers: Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods. This variable is pivotal for inter-vendor comparisons, enabling the analysis to uncover unique pricing strategies, competitive behaviors, and market segmentation. By examining differences in pricing across these suppliers, the analysis can identify which vendors consistently offer lower prices, adopt aggressive discounting practices, or align their pricing with premium branding.

Another critical variable is month, recorded as a numeric measure that captures the time of observation on a monthly scale. This variable facilitates the investigation of seasonal trends, allowing for the analysis of how strawberry prices fluctuate throughout the year. Seasonal dynamics may reflect changes in supply (e.g., peak harvest periods), demand (e.g., holidays or special occasions), and vendor-specific pricing adjustments. By integrating month with pricing data, the analysis can identify whether prices tend to peak or dip during specific times of the year and how these patterns vary among vendors.

These variables collectively provide a rich dataset for conducting both granular and aggregate analyses of strawberry pricing. The inclusion of current and old prices enables an in-depth exploration of vendor-specific pricing adjustments, while the categorical vendor variable allows for competitive benchmarking across suppliers. The month variable introduces a temporal dimension, facilitating the study of seasonal effects on pricing. Together, this dataset offers a comprehensive foundation for understanding the complex interplay between market forces, supplier behavior, and seasonal trends, yielding actionable insights for consumers seeking the best value and suppliers aiming to optimize their pricing strategies.

## Outcome variable

@@

```{r}
#| label: fig-currentpricebyvendor
#| fig-cap: "Average Current Price by Vendor"
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_parquet(here("data","02-analysis_data","analysis_data.parquet"))
ggplot(analysis_data, aes(x = vendor, y = current_price, fill = vendor)) +
  stat_summary(fun = "mean", geom = "bar", color = "black") +
  labs(
    x = "Vendor",
    y = "Average Current Price (CAD)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

```{r}
#| label: fig-monthlytrend
#| fig-cap: "Monthly Trend of Current Prices by Vendor"
#| echo: false
#| warning: false
#| message: false
ggplot(analysis_data, aes(x = month, y = current_price, group = vendor, color = vendor)) +
  stat_summary(fun = "mean", geom = "line", size = 1) +
  geom_point(stat = "summary", fun = "mean", size = 2) +
  labs(
    x = "Month",
    y = "Average Current Price (CAD)"
  ) +
  theme_minimal()
```

This graph(@fig-monthlytrend) shows the monthly trend of average current strawberry prices across different vendors over several months. `Metro` consistently exhibits the highest prices, with noticeable fluctuations, peaking in mid-year, indicating potential seasonal effects or changes in supply and demand. In contrast, budget-friendly vendors like `No Frills` and `Save-On-Foods` maintain the lowest prices, demonstrating pricing stability aimed at cost-conscious consumers. Mid-range vendors such as `Loblaws`, `T&T`, and `Voila` show relatively stable but slightly variable trends, with occasional peaks suggesting targeted pricing adjustments. Overall, the graph highlights clear segmentation among vendors, with premium vendors like `Metro` adapting dynamically to market conditions, while budget vendors maintain consistent affordability. This trend could reflect differing business strategies to cater to diverse consumer preferences.

@@

Predictor Variables: Exploring Factors Influencing Strawberry Prices

1. Vendor

The vendor variable is categorical, identifying the supplier of strawberries. Each vendor employs unique pricing strategies, leading to significant variability in pricing across the eight vendors (Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods). Premium vendors like Metro and Galleria tend to target quality-conscious consumers, reflected in their higher average prices. In contrast, budget-oriented vendors like No Frills and Save-On-Foods maintain lower prices to attract cost-sensitive shoppers. The variability in pricing, shown through boxplots, highlights how vendor identity influences price distributions and captures the strategic segmentation of the Canadian strawberry market.

2. Old and Current Prices

The dataset includes two continuous price variables: old price (original price) and current price (price after any promotions or discounts). The difference between these two variables provides insights into promotional behavior and markdown strategies. Vendors like No Frills and Save-On-Foods typically show small gaps between old price and current price, suggesting limited use of deep discounts, whereas Metro and Walmart exhibit larger gaps, reflecting more aggressive promotional activity. Boxplots comparing old price and current price by vendor demonstrate how discounts influence consumer pricing perceptions and vendor competition.

3. Month (month)

The month variable is numeric, capturing the month of observation. Seasonal patterns play a crucial role in strawberry pricing, with supply and demand fluctuations driving price variability. For instance, prices may drop during peak harvest months and increase during off-season periods when strawberries are scarce. Line charts tracking average monthly prices show clear temporal trends, with peaks and dips aligning with these seasonal dynamics. By analyzing month in combination with other predictors, this variable helps uncover how vendors adjust pricing strategies to reflect seasonal demand and supply conditions.

The interplay of vendor, old price, current price, and month allows for a more comprehensive analysis of pricing strategies. For example, vendors like Metro may use larger discounts (old price vs. current price) during peak seasons to maintain competitiveness, while budget vendors like No Frills may maintain consistent pricing throughout the year. A combined analysis using bar charts and line plots highlights how vendor strategies vary with time, providing insights into the dynamic nature of strawberry pricing.

@@

- Quantify the impact of each predictor (e.g., vendor, original price, and month) on strawberry prices.
- Analyze how the gap between original (`old_price`) and discounted (`current_price`) prices varies across vendors.
- Examine seasonal trends in strawberry prices to identify periods of high or low prices.
- Provide actionable insights into vendor pricing strategies and market dynamics.

## Model Description

@@

**Predictors in the Model**

- **Vendor (`vendor`)**: A categorical variable identifying the supplier of strawberries. Different vendors adopt distinct pricing strategies that influence strawberry prices.

- **Original Price (`old_price`)**: A continuous variable representing the original price of strawberries before discounts or promotions. This variable reflects vendor-specific baseline pricing strategies.

- **Month (`month`)**: A numeric variable capturing the observation month, enabling the analysis of seasonal trends and time-based fluctuations.

- **Price Gap (`old_price - current_price`)**: A derived variable quantifying the magnitude of discounts or markdowns applied by vendors.

**Model set-up**

@@

where

- \( y_i \): Current price (`current_price`) of strawberries for observation \( i \) (dependent variable).
- \( \beta_0 \): Intercept, representing the baseline price (reference vendor, average original price, and earliest month).
- \( \beta_{\text{vendor}[i]} \): Vendor-specific effect, capturing how each supplier’s prices deviate from the baseline vendor.
- \( \beta_{\text{old\_price}} \): Effect of the original price, representing how the initial pricing influences the current price.
- \( \beta_{\text{month}} \): Effect of the observation month, capturing seasonal or temporal trends in strawberry pricing.
- \( \epsilon_i \): Residual error, accounting for variability not explained by the predictors.

**Priors in the Bayesian Framework**

Coefficients (\( \beta \)):

@@

Residual Standard Deviation (\( \sigma \)):

@@

The priors in this analysis reflect reasonable assumptions about the relationships between predictors and strawberry prices. Coefficients $(\beta)$ are assigned normal priors $(\operatorname{Normal}(0,2.5)$ ) to reflect the expectation that most effects are small and centered around zero, with moderate variability. For example, $\beta_{\text {old_price }}$ is expected to be positive, reflecting the influence of original prices on current prices, while $\beta_{\text {month }}$ accounts for typically small seasonal pricing effects. The intercept $\left(\beta_0\right)$ also follows a Normal $(0,2.5)$, assuming the baseline price is near zero with moderate deviations. Finally, the residual standard deviation $(\sigma)$ is given an exponential prior $(\operatorname{Exponential}(1))$ to ensure positivity and account for variability not captured by the predictors.

@@

## Model justification

A Bayesian linear regression model is a suitable approach for this analysis because it balances simplicity, interpretability, and flexibility while quantifying uncertainty in parameter estimates. This model is particularly effective for exploring the continuous outcome variable, current_price, and its relationship with predictors such as vendor, month, and original price (old price). Unlike traditional linear regression, which assumes a deterministic relationship between predictors and the outcome, the Bayesian framework provides posterior distributions for all parameters, enabling richer and more nuanced inferences. This approach allows us to incorporate prior knowledge, account for uncertainty, and provide probabilistic insights into strawberry pricing.

The Bayesian framework also handles sparse or imbalanced data effectively by leveraging prior distributions to stabilize parameter estimates. For example, when certain vendors have fewer observations, the priors prevent overfitting and provide regularized estimates, making the model robust to variability in data distribution. This ensures reliable insights into vendor-specific pricing strategies, seasonal trends, and the influence of baseline (old price) on observed (current price) prices.

Benefits of Using Bayesian Regression:

1. Uncertainty Quantification:

   - The Bayesian approach provides posterior distributions for each parameter, allowing for uncertainty quantification. For example, the posterior distribution of \( \beta_{\text{vendor}} \) helps quantify the uncertainty around vendor-specific pricing effects.

2. Incorporating Prior Knowledge:

   - Priors enable the incorporation of domain knowledge into the analysis. For instance:
     - \( \beta_{\text{vendor}} \): Reflects the expectation that most vendor effects are close to zero, with some deviations.
     - \( \beta_{\text{month}} \): Captures small but meaningful seasonal effects.
     - \( \beta_{\text{old\_price}} \): Represents the assumption that higher original prices lead to higher observed prices.

3. Handling Sparse Data:

   - Bayesian models stabilize estimates when some categories (e.g., certain vendors) have fewer observations. Priors regularize the estimates, reducing the risk of overfitting.

4. Probabilistic Insights:

   - Posterior distributions provide credible intervals, making it easier to interpret parameter estimates. For instance, a 95% credible interval for \( \beta_{\text{old\_price}} \) of \([0.8, 1.2]\) means there is a 95% probability that the original price effect lies within this range.

## Choice of Predictors

1. Vendor

- **Role**: The vendor variable is categorical, capturing supplier-specific pricing strategies. Each vendor adopts distinct pricing approaches based on branding, market positioning, and target demographics.

- **Importance**: By including this variable, the model quantifies how much each vendor's prices deviate from the baseline vendor, providing insights into vendor-specific strategies and market segmentation.

2. Month

- **Role**: The numeric `month` variable captures seasonal trends in strawberry pricing. This allows the model to account for temporal effects, such as higher prices during off-seasons or lower prices during harvest periods.

- **Importance**: Including `month` helps the model identify recurring seasonal patterns, enabling a better understanding of time-based price fluctuations.

3. Original Price

- **Role**: The continuous variable `old_price` represents the baseline price before any discounts or promotions. This predictor captures the relationship between promotional activity and observed prices.

- **Importance**: Including this variable enables the model to assess how the gap between `old_price` and `current_price` varies across vendors and time, providing insights into promotional strategies and pricing adjustments.

@@

```{r}
#| label: fig-CI
#| fig-cap: "Posterior Credible Intervals for Model Parameters"
#| echo: false
#| warning: false
#| message: false
# Extract posterior draws
posterior_draws <- as.matrix(vendor_price_model)

# Generate credible interval plot for all parameters
mcmc_intervals(
  posterior_draws,
  pars = c("(Intercept)", "vendorLoblaws", "vendorMetro", "vendorNoFrills", 
           "vendorSaveOnFoods", "vendorTandT", "vendorVoila", "vendorWalmart",
           "month", "old_price"),
  prob = 0.95,  # 95% credible interval
  prob_outer = 0.95  # 95% outer credible interval
) +
  xlab("Parameter Estimate")
```

In this graph(@fig-CI), the posterior credible intervals for the Bayesian linear regression model highlight significant differences in strawberry pricing across vendors, seasonal trends, and the influence of original prices. Metro stands out with a clear positive effect, indicating its significantly higher prices compared to the reference vendor, consistent with its premium market positioning. In contrast, budget vendors like Save-On-Foods and No Frills exhibit negative effects, suggesting lower pricing strategies, while other vendors show smaller and more uncertain deviations. The positive, albeit modest, effect of `month` suggests a slight upward trend in prices over time, likely due to seasonal demand or inflation. The strong and precise relationship between `old_price` and `current_price` underscores the proportionality of promotional pricing strategies, where discounts often maintain the hierarchy of original prices. Together, these findings reveal actionable insights into vendor-specific pricing strategies, temporal trends, and the importance of original prices in shaping strawberry pricing dynamics.

```{r}
#| label: fig-posterior
#| fig-cap: "Posterior Distribution Histograms for Vendors and Other Parameters"
#| echo: false
#| warning: false
#| message: false
# Extract posterior samples
posterior_draws <- as.matrix(vendor_price_model)

# Plot posterior distributions using histograms
mcmc_hist(
  posterior_draws,
  pars = c("vendorLoblaws", "vendorMetro", "vendorNoFrills", 
           "vendorSaveOnFoods", "vendorTandT", "vendorVoila", "vendorWalmart", 
           "month", "old_price")
) +
  xlab("Parameter Estimate")
```

In this graph(@fig-posterior),the posterior distribution histograms illustrate the variability in vendor pricing strategies, temporal trends, and the influence of original prices on current strawberry pricing. Metro’s distribution is distinctly centered around a high positive value, confirming its premium pricing strategy compared to the reference vendor. In contrast, budget vendors like Save-On-Foods and No Frills show distributions predominantly in the negative range, highlighting their focus on affordability. The parameter for `month` exhibits a narrow distribution slightly above zero, indicating a modest upward trend in prices over time, likely driven by seasonal fluctuations or inflation. The tightly concentrated posterior for `old_price` underscores its strong and consistent influence on current prices, with higher original prices maintaining a proportional relationship with final prices. These results emphasize the diverse pricing strategies across vendors and the critical role of baseline prices in shaping market dynamics.
